---
title: "Sexual Dichromatism is associated with the slow life histories in passerine birds"
author: "Amy Duclaux"
date: "2025-05-08"
output: html_document
---

This script follows the analysis from Duclaux et al testing the relationship 
between life history traits and Sexual Dichromatism in passerine birds while 
accounting for potential confounding issues with latitude and body mass. 

# Upload packages

To complete this analysis we will need to use the standard `caper`, `phytools` 
and `TreeTools` packages for handling and plotting the phylogeny and the 
`MCMCglmm` to run the Bayesian phylogentic comparative models. As we also want
to account for uncertainty in the phylogeny itself so we will use the `mulTree`
package to run the each model over 100 randomly drawn trees. We can download
`mulTree` from github here (https://github.com/TGuillerme/mulTree).

```{r packages}

library(caper)

library(phytools)

library(MCMCglmm)

library(TreeTools)

if(!require(devtools)) install.packages("devtools")
library(devtools)
install_github("TGuillerme/mulTree", ref = "release")
library(mulTree)

```


# Upload Data

To run the analysis we will use data on the colour intensity fo passarines from
Cooney et. al 2022 (https://www.nature.com/articles/s41559-022-01714-1), Bird et 
al 2020 (https://conbio.onlinelibrary.wiley.com/doi/abs/10.1111/cobi.13486) and
from McQueen et al. 2019 (https://onlinelibrary.wiley.com/doi/abs/10.1111/ele.13375).

```{r load data}

bird_data <- read.csv("ADuclaux birds dichro lat fast-slow mating sys 1may25.csv")
head(bird_data)
nrow(bird_data)
#[1] 7052
```

For the phylogeny we use Jetz et al. 2012 (https://www.nature.com/articles/nature11631)

```{r load phylogeny}

phylo <- read.tree("bird_phylo.tre")
class(phylo)

```


We will log body mass, maximum lifespan and age at first reproduction and set
latitude to absolute values.

```{r clean data}
dichro_bird_data <- data.frame(animal= bird_data$animal,
                               latin_name= bird_data$latin_name,
                               order= bird_data$order,
                               diff_LociVS= bird_data$diff_LociVS,
                               mod_diff_LociVS= bird_data$mod_diff_LociVS,
                               dichro= bird_data$dichro, 
                               sexual_dichro= bird_data$sexual_dichro,
                               midpoint_lat= abs(bird_data$midpoint_lat),
                               log_mass= log10(bird_data$mass),
                               survival= bird_data$survival,
                               log_max_lifespan= log10(bird_data$max_lifespan),
                               log_age_first_breed= log10(bird_data$age_first_breed),
                               mating_system= as.factor(bird_data$mating_system_binary))

dichro_bird_data <- unique(dichro_bird_data)
dichro_bird_data <- na.omit(dichro_bird_data)

head(dichro_bird_data)

```


Create a multiphlyo object of 100 phylogenies which are matched to our dataset.


```{r match_phylo}

#Create a dataset of just the names to match
match_data <- dichro_bird_data[ ,c("latin_name", "order")]

comp_phylo_birds <- list()

for (i in 1:100){
  
comp_phylo_birds[[i]] <- comparative.data(phy = phylo[[i]], 
                                            data = match_data,
                                            names.col = latin_name, 
                                            vcv= TRUE)$phy}
class(comp_phylo_birds)
#[1] "list"

multiphylo_birds <- as.multiPhylo(comp_phylo_birds)
class(multiphylo_birds)
#[1] "multiPhylo"

```

We can now create a mulTree object that contains the 100 phylogenies and the data

```{r multree_birds}
multree_birds <- as.mulTree(data= dichro_bird_data,
                            tree= multiphylo_birds,
                            taxa= "latin_name")

```

We can now set the number of iterations, the thinning and the burnin we will use
for the models. We will also set the non-informative prior we will use througout
for the models. 

```{r priors}
parameters <- c(120000, 50, 20000)

priors <- list(R= list(V= 1, nu= 0.002), 
               G = list(G1=list(V= 1,n= 2, nu= 0.002)))

```


# Models

In order to define the slow-fast continuum we run a model for each of the life
history traits of maximum lifespan, survival and age at first reporduction as 
response variables and mass and latitude as explanatory variables. We then use 
the residuals from these models in a PCA in order to define slow fast contiuum. 

## Survial model

First we run the survival model, due to the long run time of this model we will 
not evaluate it here but read in a previous run.

```{r Survial multree, eval = FALSE}

#Define the model
survival_resids_formula <- survival ~ log_mass  + midpoint_lat

mulTree(mulTree.data = multree_birds,
        formula = survival_resids_formula,
        priors = priors,
        parameters = parameters,
        output = "bird_survival_resids",
        ESS= 1000,
        chains= 2)

```

Read in the Survial model

```{r Survial multree read}

multree_surv_res <- read.mulTree("bird_survival_resids")
str(multree_surv_res)

summary(multree_surv_res)

read.mulTree("bird_survival_resids", convergence= TRUE)


surv_resids <- multree_birds$data[,10] - c(summary(multree_surv_res)[1] 
                                           + multree_birds$data$log_mass*summary(multree_surv_res)[2]
                                           + multree_birds$data$midpoint_lat*summary(multree_surv_res)[3])


surv_resids

```


## Max lifespan

Next we run the max lifespan model, due to the long run time of this model we 
will not evaluate it here but read in a previous run.

```{r Max_lifespan multree, eval = FALSE}
max_lifespan_resids_formula <- log_max_lifespan ~ log_mass  + midpoint_lat

mulTree(mulTree.data = multree_birds,
        formula = max_lifespan_resids_formula,
        priors = priors,
        parameters = parameters,
        output = "bird_max_lifespan_resids",
        ESS= 1000,
        chains= 2)
```

Read in the max lifespan model

```{r Max_lifespan multree, eval = FALSE}

multree_max_lifespan_res <- read.mulTree("bird_max_lifespan_resids")
str(multree_max_lifespan_res)

summary(multree_max_lifespan_res)

read.mulTree("bird_max_lifespan_resids", convergence= TRUE)


max_life_resids <- multree_birds$data[,11] - c(summary(multree_max_lifespan_res)[1] 
                                           + multree_birds$data$log_mass*summary(multree_max_lifespan_res)[2]
                                            + multree_birds$data$midpoint_lat*summary(multree_max_lifespan_res)[3])
max_life_resids


```


## Age at first reproduction

Next we run the age at first reproduction model, due to the long run time of this 
model we will not evaluate it here but read in a previous run.

```{r age at first reproduction, eval = FALSE}
first_breed_resids_formula <- log_age_first_breed ~ log_mass  + midpoint_lat

mulTree(mulTree.data = multree_birds,
        formula = first_breed_resids_formula,
        priors = priors,
        parameters = parameters,
        output = "bird_first_breed_resids",
        ESS= 1000,
        chains= 2)
```

Read in the Age at first reproduction model

```{r age at first reproduction read}

multree_first_breed_res <- read.mulTree("bird_first_breed_resids")
str(multree_first_breed_res)

summary(multree_first_breed_res)

read.mulTree("bird_first_breed_resids", convergence= TRUE)


first_breed_resids <- multree_birds$data[,12] - c(summary(multree_first_breed_res)[1] 
                                           + multree_birds$data$log_mass*summary(multree_first_breed_res)[2]
                                           + multree_birds$data$midpoint_lat*summary(multree_first_breed_res)[3])
first_breed_resids


```


# PCA

We then create a PCA using the residuals form the above three models

First create a dataset of the residuals.
```{r resids dataframe}
res_fast_slow_bird_data <- data.frame(surv_resids, 
                                      max_life_resids, 
                                      first_breed_resids)

```

Next run the PCA

```{r PCA}
pca_data <- data.frame(surv_resids= res_fast_slow_bird_data$surv_resids,
                       max_life_resids= res_fast_slow_bird_data$max_life_resids,
                       first_breed_resids= res_fast_slow_bird_data$first_breed_resids)
biplot(prcomp(pca_data))
prcomp(pca_data)

#The first axis best defines the slow-fast contiuum so we will use that.
pca_data2 <- prcomp(pca_data)
fast_slow_cont <- pca_data2$x[,1]


```

# Main model

We now add the fast-slow data to the multree data to set up for the main model.

```{r final model set up}
fast_slow_bird_data <- data.frame(multree_birds$data, fast_slow_cont)
head(fast_slow_bird_data)
nrow(fast_slow_bird_data)
#[1] 2371

fast_slow_bird_data <- data.frame(animal= fast_slow_bird_data$animal,
                                  latin_name= fast_slow_bird_data$sp.col,
                                  diff_LociVS= fast_slow_bird_data$diff_LociVS,
                                  mod_diff_LociVS= fast_slow_bird_data$mod_diff_LociVS,
                                  dichro= fast_slow_bird_data$dichro, 
                                  sexual_dichro= fast_slow_bird_data$sexual_dichro,
                                  midpoint_lat= fast_slow_bird_data$midpoint_lat,
                                  log_mass= fast_slow_bird_data$log_mass,
                                  mating_system= fast_slow_bird_data$mating_system,
                                  fast_slow_cont= fast_slow_bird_data$fast_slow_cont)
                               
multree_birds_fast_slow <- as.mulTree(data= fast_slow_bird_data,
                                      tree= multiphylo_birds,
                                      taxa= "latin_name")
```

Running multree model with diff_LociVS residual fast_slow_cont, midpoint_latitude, 
log10(mass) and mating_system. As its a long run we will read in a previous run.

```{r Main multree model, eval = FALSE}

multree_fsc_diff_LociVS <- diff_LociVS ~ fast_slow_cont + midpoint_lat 
+ mating_system + log_mass

mulTree(mulTree.data = multree_birds_fast_slow,
        formula = multree_fsc_diff_LociVS,
        priors = priors,
        parameters = parameters,
        output = "fsc_diff_LociVS_June_4",
        ESS= 1000,
        chains= 2)
```

Read in a previous run of the model 

```{r Main multree model read}

multree_fsc_diff_LociVS_model <- read.mulTree("fsc_diff_LociVS")

summary(multree_fsc_diff_LociVS_model, use.hdr = FALSE)

```
